Automatically Splitting a Staged Lambda Calculus

### 2015-07-09 paper/relatedwork.tex
Clean up the graphics related work section to be as factual as possible.

In the partial evaluation community, stage splitting has been investigated 
mostly as a preprocessing step for self-applied partial evaluators. 
For instance, \cite{Mogensen89a} and \cite{DeNiel91} defined splitting algorithms
for programs consistening of a set of first-order recursive functions.
Those works use a grammar-based binding-time specification that's 
relatively complicated compared to the type system of \lang.
This complication mostly arises from the decision to start with an untyped language:
one must first find a way to say ``argument $x$ will be a tuple'' before one can say
``the first component of $x$ is available at the second stage,'' 
and so much of their binding-time specification just amounts to a type system.
Formulations more like \lang\ (namely the two-level system of \cite{NN92-twolevel}) were known at the time,
but \cite{Mogensen89a} found these unsatisfactory because they could not describe data structures with mixed stage.
Davies essentially solved this limitation in \cite{davies96} (pending our further tweak with $\curr$),
and the question of stage splitting as we have presented it has been ripe ever since.
Binding time analysis for a static higher-order language was visited in \cite{Mogensen89b},
but the implications for splitting were not explicitly considered.

Splitting-like algorithms have also been a topic of interest in
programming systems for computer graphics, where to achieve
high performance, programs are typically manually separated into
components by frequency of execution. (Indeed, separation into a pre-determined
set of per-frequency modules is \emph{required} by the execution model
of modern graphics architectures.)  
The software engineering challenges of modifying programs in the split form (see \cite{Foley:2011}) have led to
suggestions of defering the task of splitting to the compiler.
Early work \cite{knoblock96} sought to determine the staging of programs automatically through use of binding time analysis,
however later work \cite{Proudfoot:2001,Foley:2011,He:2014} has focused on giving programmers explicit control of the staging through 
programming language features. 
All of these efforts in graphics have been limited to simple, imperative languages.

%% \section{Related Work}

%% \ur{
%% \begin{itemize}
%% \item 
%% I find it hard to distinguish between blue and black on printed paper.
%% \item
%% $\pause$ should be called ``promote'' or something like that. I
%%   believe this is what Taha calls cross stage persistence.  THat work should
%%   be cited.
%% \end{itemize}
%% }

%% \crem{\emph{This text came from section 3:}
%% \lang\ is adapted from \cite{davies96}, and is very similar to other staged
%% languages considered in the partial evaluation and metaprogramming literature.
%% Unlike \cite{davies96}, we restrict \lang\ to exactly two stages, to simplify
%% our presentation of our stage-splitting algorithm. We expect that the techniques
%% presented here extend to more stages.
%% Our dynamic semantics for \lang\ differ from those given in \cite{davies96} and
%% the metaprogramming literature, because we avoid duplicating stage \bbtwo\
%% computations. In addition to being more efficient, our non-duplicating semantics
%% are easier to compare with our splitting algorithm, whose development and
%% exposition is our main goal.
%% }

%% There are several distinctive aspect of our work which we wish to compare and contrast.
%% Firstly, \lang\ is a simply typed lambda calculus 
%% with explicit stage annotations and a non-duplicating semantics.
%% Its statics and dynamics enforce no backward information flow,
%% which a is necessary property to define a splitting transformation
%% that is equivalent to the dynamics.
%% Finally, the splitting transformation itself operates on any well-typed open term in \lang.

%% Partial evaluation is a well-studied technique for specializing programs to some of their input.  
%% Often, the goal of partial evaluation systems is to go all the way from unstaged code to a specialized version.
%% Although this problem definition implicitly includes binding time analysis,
%% many presentations factor out that step and start with a stage-annotated representation, as we have done.
%% We've shown that our dynamic semantics can be used to perform partial evaluation 
%% [need to do more reading to know exactly how it compares].

%% One unifying feature of partial evaluation systems is that they operate on code which is closed at the first stage,
%% and so they must necessarily unfold recursive calls and wait for all stage \bbone\ input to operate.
%% There has been some work to avoid these issues by templating [need citations].
%% Stage-splitting, in contrast, goes all the way to operating on terms which are open on first stage variables,
%% and so these problems are avoided.

%% Stage-splitting itself has appeared in the literature---under 
%% the names {\em pass separation} and {\em data specialization}--- for various simpler input languages.

%% Most often, splitting is considered a compiler optimization (\cite{jorring86},\cite{knoblock96})
%% to minimize recomputation.
%% In this use case, only the arguments to the program are annotated with stages, 
%% and the splitter is left to perform an {\em ad hoc} binding time analysis to determine the 
%% internal staging of the program.
%% Because this occurs without programmer supervision, the binding time analyzers for these uses are 
%% reluctant to place stage \bbone\ code within stage \bbtwo\ conditionals, 
%% since that would result in speculative behavior that could increase the overall runtime.
%% We don't worry about this since the annotations of \lang\ are assumed to represent the will of a programmer.

%% Like the previous example, the Spark language (\cite{sparkThesis}) uses staging to minimize recomputation in real-time rendering applications.  
%% But instead of using a binding-time analysis to determine what parts of the program to put in what shader, 
%% Spark gives the programmer manual control of term staging using a language that is in spirit similar to \lang.

%% We are also not the first to use stage splitting to derive algorithms.
%% \cite{malmkjaer89}, expanding on the work of \cite{barzdins88}, defined a stage-splitter
%% for a language of tail-recursive first-order equations with explicit staging annotations [double check that],
%% and considered applications to string matching.

%% Our work is distinct from all previous stage splitters in the following ways:
%% \begin{itemize}
%% \item Our input language, \lang, has a disciplined type theory.
%% \item \lang\ has full first-class functions (and by extension, recursion), as well as sum types.
%% \item Our stage splitting is proven correct.
%% \end{itemize}

%% The goal of {\em metaprogramming} (\cite{taha-thesis-99}, \cite{devito13}, \cite{davies01}) is 
%% to use code in a host language to write code in an object language.
%% Like partial evaluation and stage-splitting, this can be modeled by a staged language with explicit annotations.
%% Instead of identifying stages with points in time, stages are identified with host/object levels of the language,
%% and so type safety ensures the proper separation of levels of the language and safety of the constructed object code.

%% But the differences in goals of metaprogramming do cause a few differences in form.
%% For example, it's generally preferred in the metaprogramming context to use 
%% a dynamics that duplicates the contents of $\next$ blocks.
%% Moreover, in metaprogramming systems it makes sense to define an operator, \texttt{run}, 
%% which evaluates stage \bbtwo\ code to get an answer back at stage \bbone.
%% This is essentially a form a backwards information flow.
%% Thus, the \texttt{run} feature is directly incompatible with a temporal interpretation of stages,
%% which stage-splitting requires.

%% \cite{davies96} explored the connection between linear temporal logic and its corresponding type system (which is nearly identical to that for \lang), 
%% and showed the equivalence between that type system and existing one for binding time analysis. 
%% That work also provided a metaprogramming-style duplicating dynamic semantics, which we purposely avoided.
%% That said, the dynamics of \cite{davies96} is otherwise very similar to ours.
%% In fact, if we change the $\next$ and $\prev$ rules of ours to 
%% \begin{mathpar}
%% \infer {\diaone {\next~e}{\cdot,\next~q}} {\diatwo e q} \and
%% \infer {\diatwo{\prev~e}{v}} {\diaone e {\cdot,\next~v}} 
%% \end{mathpar}
%% and treat $\next~q$ as a partial value for any residual $q$,
%% then our semantics becomes rule-for-rule isomorphic to that from \cite{davies96}. This essentially bypasses the
%% environment bookkeeping in $\redonesym$, by inlining residuals instead of
%% hoisting them in \verb|let|-bindings.
%% From this it's clear that the semantics of \cite{davies96} and ours always produce the same value when they both terminate.
%% {\em However because \lang's semantics dictate that a reified residual will always be evaluated, regardless of whether its result is consumed, \lang\ programs terminate strictly less often than that of \cite{davies96}}.

@@

%% \paragraph{Supercompilation.}
%% ~\cite{Turchin86}

%% \paragraph{Intro.}

%% A predominating theme in the design, implementatation, and
%% optimization of algorithm and software is frequency reduction and
%% precomputation.  The idea is to reduce the frequency of computations
%% by reducing their frequency and ideally by precomputing when possible.

%% The reason for why this theme is as predominant as it is that it is
%% naturally common in many actual applications.  For example, we may
%% have a collection of names or numbers that we frequently lookup, as
%% for example part of a membership test as part of a larger system.  In
%% some applications areas such as in graphics (the motivation for our
%% work), this is so common that there is a relatively large literature
%% on speeding up graphics computations by taking advantage of frequency
%% reduction and precomputation.

%% One way to express frequency of computations is to allow the
%% programmer to stage computation explicitly by making explicit the
%% stage of each computation and then by moving computations between
%% stages to improve efficiency by performing work as early (to save time
%% later when performance might be more critical) and  as infrequently
%% as possible (to reduce overall time).

%% For example, in meta-programming~\cite{}, the programmer can assign a
%% program expression a stage (stage 1, 2, 3, ...), which indicates the
%% stage at which that expression evaluates.  Combined with the ability
%% for earlier stages to construct expressions of later stages,
%% meta-programming techniques can be used to construct efficient
%% programs that improve efficiency over their non-staged counterparts by
%% reducing frequency reduction and precomputation.  For example, the
%% programmer can write a program that takes a list of numbers $\ell$ as
%% a first stage value and returns a second-stage function that performs
%% a fast membership test on that list $\ell$ by inlining all the
%% comparisons.  When given a key to lookup in the second stage, the
%% membership test function would quickly return the result.

>>>>

In the partial evaluation community, stage splitting has been investigated 
mostly as a preprocessing step for self-applied partial evaluators. 
For instance,~\cite{Mogensen89a} and~\cite{DeNiel91} defined splitting algorithms
for programs consistening of a set of first-order recursive functions.
Those works use a grammar-based binding-time specification that's 
relatively complicated compared to the type system of \lang.
This complication mostly arises from the decision to start with an untyped language:
one must first find a way to say ``argument $x$ will be a tuple'' before one can say
``the first component of $x$ is available at the second stage,'' 
and so much of their binding-time specification just amounts to a type system.
Formulations more like \lang\ (namely the two-level system of~\cite{NN92-twolevel}) were known at the time,
but~\cite{Mogensen89a} found these unsatisfactory because they could not describe data structures with mixed stage.
Davies essentially solved this limitation in~\cite{davies96} (pending our further tweak with $\curr$),
and the question of stage splitting as we have presented it has been ripe ever since.
Binding time analysis for a static higher-order language was visited in \cite{Mogensen89b},
but the implications for splitting were not explicitly considered.

Splitting algorithms have also been a topic of interest in programming
systems for computer graphics, where to achieve high performance,
programs are typically manually separated into components by frequency
of execution. (Indeed, separation into a pre-determined set of
per-frequency modules: e.g., per frame, per pixel, per triangle, is
\emph{required} by the execution model of modern graphics
architectures.)  The software engineering challenges of modifying
graphics programs written in a split form have led to explorations of
deferring the task of splitting to a compiler\,\cite{Foley:2011}.
While early work sought to determine the staging of programs
automatically via binding time analysis\,\cite{knoblock96}, more
recent designs, like \lang, require explicit staging by the
programmer\,\cite{Proudfoot:2001,Foley:2011,He:2014}.  However,
splitting algorithms developed in all of these efforts have been
limited to simple, imperative graphics shading languages that lack
simple features such as functions.

@@
